# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

Reading in the dataset

```{r echo=TRUE}
activity <- read.csv(unz("activity.zip", "activity.csv"))
```

Quick look at the dataset.

```{r echo=TRUE}
names(activity)
summary(activity)
```


A histogram of the total number of steps taken each day.

```{r echo=TRUE}
day.counts <- aggregate(steps ~ date, data= activity, FUN=sum)
hist(day.counts$steps, xlab="No. steps", ylab="No. days", 
     main="Histogram of total number of steps taken each day")
```


## What is mean total number of steps taken per day?

The mean number of steps taken each day.

```{r echo=TRUE}
mean(day.counts$steps)
```

The median number of steps taken each day.

```{r echo=TRUE}
median(day.counts$steps)
```



## What is the average daily activity pattern?

Time series plot of the average number of steps taken (averaged across all days) versus the 5-minute intervals

```{r echo=TRUE}
mean.steps.by.interval <- aggregate(steps ~ interval, data= activity, FUN=mean)
plot(mean.steps.by.interval$interval, mean.steps.by.interval$steps, type="l",xlab="Interval", ylab="Avg steps (all days)")
```

## Imputing missing values

Calculate and report the total number of missing values in the dataset

```{r echo=TRUE}
num.missing <- sum(is.na(activity))
```

Create a new dataset that is equal to the original dataset with the missing data filled in  


```{r echo=TRUE}
activity.imputed <- activity
rownames(mean.steps.by.interval) <- mean.steps.by.interval$interval

activity.imputed[is.na(activity.imputed),]$steps <- mean.steps.by.interval[as.character(activity.imputed[which(is.na(activity.imputed)),]$interval),]$steps

sum(is.na(activity.imputed))
```

Histogram of the total number of steps taken each day


```{r echo=TRUE}
day.counts.imputed <- aggregate(steps ~ date, data= activity.imputed, FUN=sum)
hist(day.counts2$steps, xlab="Number of steps", ylab="Number of days", main="Histogram of total number of steps taken each day")
```

Calculate and report the mean and median total number of steps taken per day.

The mean number of steps taken each day.

```{r echo=TRUE}
mean(day.counts.imputed$steps)
```

The median number of steps taken each day.

```{r echo=TRUE}
median(day.counts.imputed$steps)
```

Do these values differ from the estimates from the first part of the assignment?

```{r echo=TRUE}
mean(day.counts.imputed$steps) - mean(day.counts$steps)
```
No difference between the two mean values.

```{r echo=TRUE}
median(day.counts.imputed$steps) -median(day.counts$steps)
```

There is only a small difference between the two median values. The median value for the dataset with imputed missing data increased the median but about 1 step.

What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r echo=TRUE}
sum(day.counts.imputed$steps) - sum(day.counts$steps)
```

The total daily number of steps over the days goes up by over 86000 steps with imputed missing data.


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or a weekend day.

```{r echo=TRUE}
# set date as a date type, so it can be used with weekdays function
activity.imputed$date <- as.Date(activity.imputed$date) 
# initialise as everything is weekend (yeah, we wish)
activity.imputed$weekday <- "weekend"
# set Monday - Friday as weekday
activity.imputed[weekdays(activity.imputed$date) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),]$weekday <- "weekday"
# set levels
levels(activity.imputed$weekday) <- c("weekday", "weekend")
```

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
# weekday means
mean.steps.by.interval.weekday <- aggregate(steps ~ interval, data=activity.imputed[activity.imputed$weekday=="weekday",], FUN=mean)
mean.steps.by.interval.weekday$weekday <- "weekday"
# weekend means
mean.steps.by.interval.weekend <- aggregate(steps ~ interval, data=activity.imputed[activity.imputed$weekday=="weekend",], FUN=mean)
mean.steps.by.interval.weekend$weekday <- "weekend"
# combine by rows
mean.steps <- rbind(mean.steps.by.interval.weekday, mean.steps.by.interval.weekend)
# set levels
levels(mean.steps$weekday) <- c("weekday", "weekend")
# plot
library(lattice)
attach(mean.steps)
xyplot(steps ~ interval | weekday, type = 'l', xlab = "Interval", ylab = "Number of steps", layout = c(1, 2))

```

From the plot we can see there are differences in activity patterns between weekdays and weekends. On average, weekdays have more activity (steps), especially in the mid-to-late morning intervals, with weekends looking like activity starts later in the day. However, weekends appear to involve higher levels of activity between mid-morning and evening time periods, with more steps on average taken compared to lower levels of activity on weekdays in the afternoon.